probKanren: A Simple Probabilistic extension for microKanren

### 2021-09-18 paper/main.tex
Misc fixes

But the conceptual simplicity of this sort of generalisation is not
reflected in the complexity of many existing probabilistic logic
programming systems.  They often involve making implementing
sophisticated inference algorithms and the underlying systems are not
just implemented on top of existing logic programming systems.

We believe that a conceptually simple extension deserves a
conceptually simple implementation to go along with it. We thus
contribute a simple way to extend microkanren a small logic
programming domain-specific language such that it becomes a
probabilistic programming language.

@@

\textsc{microKanren}\cite{10.1145/2989225.2989230, daniel2018reasoned}
is a pure logic programming language embedded in Scheme. The language
consists of a set of terms, a set of goal primitives, and two run
functions that execute queries in the language. The goal primitives
consist of a \texttt{fresh} form for introducing logic variables, a
unification primitive \texttt{==}, a conjunction combinator
\texttt{conj}, a disjunction combinator \texttt{disj}, and ways to
define and apply relations. Further forms are shown in detail in
\Cref{fig:grammar}.

Goal expressions represent a possibly backtracking search
procedure. These goals all take as input some state and output a
stream of possible output states. We run these goals by starting with
an initial state that holds an empty substitution dictionary, and
passing it into the top-level goal expression which then passes it
recursively down to sub-expressions. Encountering \texttt{fresh} adds
a new logical variable to the substitution; \texttt{conj} will apply
the first goal to the state passed in, and then apply the second goal
to the states associated with each resulting stream from the first
goal and finally unify the results; \texttt{disj} will apply both
goals to the same input state and concatenate the resulting streams;
and \texttt{==} unifies its arguments in the context of the current
state, discarding any streams which fail to unify.

@@

To create \textsc{probKanren}, we extend this grammar with distribution
clauses such as \texttt{normal} and \texttt{bern}. Distribution clauses
take as arguments the parameters of the distribution and a last argument
representing a draw from that distribution. For example, in \texttt{(normal 0 1 x)}
means \texttt{x} represents a draw from the $\mathcal{N}(0,1)$ distribution.

@@

We follow the semantics of \cite{staton2016semantics}, where each
language form has the denotational semantics that a
measurable function, and an operational semantics of a sampler.

@@

There are many ways to inference algorithms for probabilistic
programming languages, but methods based on likelihood-weighting and
Sequential Monte Carlo algorithms are the simplest to implement.

@@

We know in $T$ rounds we compute the desired distribution as follows:

@@

This extension does not modify the search and streams are managed
exactly as in \textsc{microKanren}. An additional advantage, due to
the completeness of the \text{microKanren} search being complete, is
that we are guaranteed to recover the true posterior as all paths of
the search space will eventually be explored given enough particles.

@@

We made a simple to implement extension to \textsc{microKanren} that
lets us support probabilistic inference on both discrete and
continuous distributions. The approach does not require modifying the
underlying search algorithm or touching any of the backtracking code and
comes with a theoretical guarantee that if the underlying search is
complete then the probabilistic extension will recover the true
posterior given enough particles.

>>>>

But the conceptual simplicity of this sort of generalisation is not
reflected in the complexity of many existing probabilistic logic
programming systems.  They often involve implementing sophisticated
inference algorithms and the underlying systems are not just
implemented on top of existing logic programming systems.

We believe that a conceptually simple extension deserves a
conceptually simple implementation to go along with it. We thus
contribute a simple way to extend \textsc{microKanren}, a small logic
programming domain-specific language such that it becomes a
probabilistic programming language.

@@

\textsc{microKanren}\cite{10.1145/2989225.2989230, daniel2018reasoned}
is a pure logic programming language embedded in Scheme. The language
consists of a set of terms, a set of goal primitives, and two run
forms that turn goal queries into a set of answers. The goal primitives
consist of a \texttt{fresh} form for introducing logic variables, a
unification primitive \texttt{==}, a conjunction combinator
\texttt{conj}, a disjunction combinator \texttt{disj}, and ways to
define and apply relations. Further forms are shown in detail in
\Cref{fig:grammar}.

Goal expressions represent a possibly backtracking search
procedure. These goals all take as input some state and output a
stream of possible output states. We run these goals by starting with
an initial state that holds an empty substitution dictionary, and
passing it into the top-level goal expression which then passes it
recursively down to sub-expressions. Encountering \texttt{fresh}
extends the lexical environment with a binding between a lexical
variable and a logic variable; \texttt{conj} will apply the first goal
to the state passed in, and then apply the second goal to the states
associated with each resulting stream from the first goal and finally
unify the results; \texttt{disj} will apply both goals to the same
input state and concatenate the resulting streams; and \texttt{==}
unifies its arguments in the context of the current state, discarding
any streams which fail to unify.

@@

To create \textsc{probKanren}, we extend this grammar with distribution
clauses such as \texttt{normal} and \texttt{bern}. Distribution clauses
take as arguments the parameters of the distribution and a last argument
representing a draw from that distribution. For example, \texttt{(normal 0 1 x)}
means \texttt{x} represents a draw from the $\mathcal{N}(0,1)$ distribution.

@@

We follow the semantics of \cite{staton2016semantics}, this is a
denotational semantics where each language form is associated with a
measurable function, and an operational semantics of a sampler.

@@

There are many inference algorithms for probabilistic programming
languages, but methods based on likelihood-weighting and Sequential
Monte Carlo algorithms are the simplest to implement.

@@

we know in $T$ rounds we compute the desired distribution as follows:

@@

This extension does not modify the search and streams are managed
exactly as in \textsc{microKanren}. An additional advantage, due to
the \text{microKanren} search being complete, we are guaranteed to
recover the true posterior as all paths of the search space will
eventually be explored given enough particles.

@@

We made a simple-to-implement extension to \textsc{microKanren} that
lets us support probabilistic inference on both discrete and
continuous distributions. The approach does not require modifying the
underlying search algorithm or touching any of the backtracking code and
comes with a theoretical guarantee that if the underlying search is
complete then the probabilistic extension will recover the true
posterior given enough particles.
